---
title: "assignment4"
author: "Florence Marti"
date: "2025-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Volumes/green_groups_exh_public/!People/Florence (EXH)/05_Courses/EpiGenetics/week4")
```

```{r, warning=FALSE, include=TRUE}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(rtracklayer)
  library(epiwraps)
  library(GenomicRanges)
  library(R.utils) #for function gunzip
  library(MASS)
})
ah <- AnnotationHub()
```

# 1. From Encode find and download the peaks for the following mouse embryonic stem cells:

## 1.a) Transcription factor p300
```{r}
download.file("https://www.encodeproject.org/files/ENCFF653YVR/@@download/ENCFF653YVR.bed.gz", destfile="p300.bed.gz")
gunzip("p300.bed.gz")
##import the peaks
p300 <- rtracklayer::import("p300.bed", format="narrowPeak")
head(p300)
```


## 1.b) Histone modification H3K4me3
```{r}
download.file("https://www.encodeproject.org/files/ENCFF974BMC/@@download/ENCFF974BMC.bed.gz", destfile="H3K4me3.bed.gz")
gunzip("H3K4me3.bed.gz")
##import peaks
H3K4me3 <- rtracklayer::import("H3K4me3.bed", format ="narrowPeak")
head(H3K4me3)
```


## 1.c) Histone modification H3K4me1

```{r}
download.file("https://www.encodeproject.org/files/ENCFF426IIV/@@download/ENCFF426IIV.bed.gz", destfile="H3K4me1.bed.gz")
gunzip("H3K4me1.bed.gz")
##import peaks
H3K4me1 <- rtracklayer::import("H3K4me1.bed", format ="narrowPeak")
head(H3K4me1)
```


## 1.d) Histone modification H3K27ac

```{r}
download.file("https://www.encodeproject.org/files/ENCFF212HAG/@@download/ENCFF212HAG.bed.gz", destfile="H3K27ac.bed.gz")
gunzip("H3K27ac.bed.gz")
##import peaks
H3K27ac <- rtracklayer::import("H3K27ac.bed", format ="narrowPeak")
head(H3K27ac)
```


## 1.e) Histone modification H3K27me3

```{r}
download.file("https://www.encodeproject.org/files/ENCFF558YNC/@@download/ENCFF558YNC.bed.gz", destfile="H3K27me3.bed.gz")
gunzip("H3K27me3.bed.gz")
##import peaks
H3K27me3 <- rtracklayer::import("H3K27me3.bed", format ="narrowPeak")
head(H3K27me3)
```


# 2. Of the p300 peaks, what proportion overlaps each of the histone modifications?
How many peaks are there in the p300 file?
```{r}
peaks_p300 <-length(p300)
peaks_p300
```

## 2.a) Overlap histone modification H3K4me3

```{r}
peaksOverlapsH3K4me3 <- overlapsAny(p300, H3K4me3, type = "any")
sumH3K4me3 <- sum(peaksOverlapsH3K4me3)
proportion_H3K4me3 <- round((100/peaks_p300*sumH3K4me3), 3)

sumH3K4me3
print(paste(proportion_H3K4me3, "%"))

```


## 2.b) Overlap histone modification H3K4me1

```{r}
peaksOverlapsH3K4me1 <- overlapsAny(p300, H3K4me1, type = "any")
sumH3K4me1 <- sum(peaksOverlapsH3K4me1)
proportion_H3K4me1 <- round((100/peaks_p300*sumH3K4me1), 3)

sumH3K4me1
print(paste(proportion_H3K4me1, "%"))
```


## 2.c) Overlap histone modification H3K27ac

```{r}
peaksOverlapsH3K27ac <- overlapsAny(p300, H3K27ac, type = "any")
sumH3K27ac <- sum(peaksOverlapsH3K27ac)
proportion_H3K27ac <- round((100/peaks_p300*sumH3K27ac), 3)

sumH3K27ac
print(paste(proportion_H3K27ac, "%"))
```


## 2.d) Overlap histone modification H3K27me3

```{r}
peaksOverlapsH3K27me3 <- overlapsAny(p300, H3K27me3, type = "any")
sumH3K27me3 <- sum(peaksOverlapsH3K27me3)
proportion_H3K27me3 <- round((100/peaks_p300*sumH3K27me3), 3)

sumH3K27me3
print(paste(proportion_H3K27me3, "%"))
```

